Ecological convergence via parallel speciation

### 2018-04-01 code/main.tex

Parallel speciation, the repeated evolution of the same modes of reproductive isolation, has fascinated evolutionary biologists because it provides strong evidence for the role of natural selection in genesis of new species (Futuyma 1986, Schluter and Nagel 1995, Schluter 2000).  In cases of parallel speciation there is often evidence of repeated evolution of ecologically important traits associated with resource use (e.g. Williams et al. 1972, Losos et al. 1998, Matthews et al. 2010, Elmer et al. 2014).  Convergent evolution in traits associated with resource use could cascade to alter community structure, creating the potential for community convergence.  Here we use the replicated ecological speciation observed in threespine stickleback fish to assess whether parallel speciation leads to community convergence in their prey.  A field study and mesocosm experiment combine to illustrate that stickleback speciation leads to convergence in the zooplankton and aquatic larval insect communities.  For example, zooplankton biomass showed a 13-fold reduction in natural lakes where speciation has occurred.  This convergence is also measurable across the aquatic-terrestrial boundary as we observed a 24\% increase in insect emergence associated with speciation.  This convergence demonstrates the profound impacts that evolutionary change can have on communities in a natural ecosystem and hints at a potential role of eco-evolutionary dynamics in ecological speciation.

The process of ecological speciation, whereby reproductive isolation between populations evolves as a by product of local adaptation, is the primary mechanism by which natural selection generates species diversity.  Yet current models of ecological speciation largely do not account for reciprocal interactions between evolutionary change and ecology, limiting our understanding of the speciation process.  Here we present the results of a two species consumer-resource model that illustrates how evolution in consumer species can drive evolution in resource species that may then feedback to dictate divergence between consumer species.  We pair this model with a large-scale observational field study and an evolutionarily replicated mesocosm experiment to assess whether ecological speciation in consumers alters resource dynamics in nature.  As predicted by the model, we found that divergence between consumer taxa alters community structure and stability in natural ecosystems, even across ecosystem boundaries.  As a whole, our study provides evidence that the process of ecological speciation can have profound effects on natural communities and suggests a currently under appreciated role for eco-evolutionary dynamics in the process of ecological speciation.

notes:
speciation as a driver of diversity and the effects of diversity on ecology
specialization as a process that can reduce stability and potentially

Intro
diversity-stability and effects of speciation on ecology

paragraph on ecological speciation, how it tends to be associated with specialization, and utility of repeated speciation

paragraph stickleback biology, cases of repeated evolution leading to repeated community structure as powerful evidence for determinism in ecology, and suitability for this question

paragraph where we lay out hypotheses using morphology and model. Model for resource dyanmics and stability.

@@

% Paragraph outlining prior results showing that biodiversity affects ecosystem function (e.g. more efficient resource use) and community stability. However, the effects of the processes that actually generate biodiversity (e.g. character displacement and the origin of species) on ecosystem function and community stability have only begun to be explored, with no clear evidence from a natural ecosystem.

@@

\begin{figure}[ht]
\includegraphics[scale = 0.6]{fig_theory.pdf}
\caption{Coevolution of competitors destabilizes a model food web. In the presence of a competitor, $C_2$, each consumer specializes on the resource it is pre-adapted to, whereas in the absence of competition, a consumer will evolve a generalist phenotype that can more equally exploit both resources (A). This ecological character displacement causes consumers to evolve a higher effective attack rate (B), which suppresses the density of both resources relative to the food web with no competing consumer (C). This enhanced resource exploitation with coevolving competitors results in a relatively less stable food web (D). The grey line in (D), where $-1\times \text{Re}(\lambda_{max})=0$, corresponds to the local stability boundary of the food web. Non-evolving model parameters for this simulation: $r=1$, $K=4$, $e=0.8$, $h=0.4$, $w=0.6$, $m=1$. Initial evolving parameters and species densities when "Evolutionary time = 0": $a_{ii}=1.93$, $a_{ij}=1.2$; $R_1=R_2=2$, $C_1=C_2=1$.}
\label{Fig:theory}
\end{figure}

\begin{figure}[ht]
\includegraphics[trim=1cm 8cm 1cm 8cm, scale=0.8]{repeat_Fig1_161109.pdf}
\caption{Panels A and B illustrate the effects of stickleback speciation on log zooplankton biomass and average size from the mesocosm experiment.  Small triangles show the mean of each independently evolved species pair of stickleback.  Panels C and D show the effects of stickleback speciation on log zooplankton biomass and average size across the year from the field study.}
\label{Fig:Zoop}
\end{figure}

@@

\begin{figure}[ht]
%\includegraphics[scale = 0.3]{meso_EmergedLogAbun_160119.pdf}
%\includegraphics[scale = 0.3]{Field_Insect_Emerge_170403.pdf}
\includegraphics[scale = 0.4]{Field_Bat_170403.pdf}
\caption{Panel A shows the number of insects emerging from the aquatic environment in the mesocosm experiment.  Panel B shows the difference in insect emergence in observational field data between paired lakes, with negative values indicating greater emergence in the species pair lakes.  Panel C shows the difference in microchiropteran foraging activity between paired lakes, with negative values indicating greater foraging above species pair lakes.}
\label{Fig:AnotherFigure}
\end{figure}

>>>>

Parallel speciation, the repeated evolution of the same modes of reproductive isolation, has fascinated evolutionary biologists because it provides strong evidence for the role of natural selection in genesis of new species (Futuyma 1986, Schluter and Nagel 1995, Schluter 2000).  In cases of parallel speciation there is often evidence of repeated evolution of ecologically important traits associated with resource use (e.g. Williams et al. 1972, Losos et al. 1998, Matthews et al. 2010, Elmer et al. 2014).  Convergent evolution in traits associated with resource use could cascade to alter community structure, creating the potential for community convergence.  Here we use the replicated ecological speciation observed in threespine stickleback fish to assess whether parallel speciation leads to community convergence in their prey.  A field study and mesocosm experiment combine to illustrate that stickleback speciation leads to convergence in the zooplankton and aquatic larval insect communities.  For example, zooplankton biomass showed a 13-fold reduction in natural lakes where speciation has occurred.  This convergence is also measurable across the aquatic-terrestrial boundary as we observed a 24\% increase in insect emergence associated with speciation.  This convergence demonstrates the profound impacts that evolutionary change can have on communities in a natural ecosystem and hints at a potential role of eco-evolutionary dynamics in ecological speciation. TEST

@@

@@

\begin{figure}[h!]
\includegraphics[scale = 0.6]{fig_theory.pdf}
\caption{Coevolution of competitors destabilizes a model food web. In the presence of a competitor, $C_2$, each consumer specializes on the resource it is pre-adapted to, whereas in the absence of competition, a consumer will evolve a generalist phenotype that can more equally exploit both resources (A). This ecological character displacement causes consumers to evolve a higher effective attack rate (B), which suppresses the density of both resources relative to the food web with no competing consumer (C). This enhanced resource exploitation with coevolving competitors results in a relatively less stable food web (D). The grey line in (D), where $-1\times \text{Re}(\lambda_{max})=0$, corresponds to the local stability boundary of the food web. Non-evolving model parameters for this simulation: $r=1$, $K=4$, $e=0.8$, $h=0.4$, $w=0.6$, $m=1$. Initial evolving parameters and species densities when "Evolutionary time = 0": $a_{ii}=1.93$, $a_{ij}=1.2$; $R_1=R_2=2$, $C_1=C_2=1$.}
\label{Fig:theory}
\end{figure}

\begin{figure}[h!]
\includegraphics[trim=1cm 8cm 1cm 8cm, scale=0.8]{repeat_Fig1_161109.pdf}
\caption{Panels A and B illustrate the effects of stickleback speciation on log zooplankton biomass and average size from the mesocosm experiment.  Small triangles show the mean of each independently evolved species pair of stickleback.  Panels C and D show the effects of stickleback speciation on log zooplankton biomass and average size across the year from the field study.}
\label{Fig:Zoop}
\end{figure}

@@

\begin{figure}[h!]
%\includegraphics{elegance}
\caption{Panel A shows the number of insects emerging from the aquatic environment in the mesocosm experiment.  Panel B shows the difference in insect emergence in observational field data between paired lakes, with negative values indicating greater emergence in the species pair lakes.  Panel C shows the difference in microchiropteran foraging activity between paired lakes, with negative values indicating greater foraging above species pair lakes.}
\label{Fig:AnotherFigure}
\end{figure}
